Report

### 2015-04-14 Lattice Reduction/Report April 2015/Report_draft.tex
First draft of April report

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[12pt,a4paper,final]{article}
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
\usepackage[latin1]{inputenc}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage{graphicx}
%\usepackage[pdftex]{graphicx}
\usepackage{psfrag}
\usepackage{color}
\usepackage{epsfig}
\usepackage{appendix}
\usepackage{caption}
\usepackage{cite}
\usepackage{ifpdf}
\usepackage[cmex10]{amsmath}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{stfloats}
\usepackage{url}
\usepackage{fixltx2e}
\usepackage{setspace} 
\usepackage{multirow}

@@

%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Parallel Acceleration of Fixed Complexity Sphere Decoder for Large MIMO Uplink System based on GPU}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Tianpei Chen}
%\IEEEauthorblockA{ID: 25504946\\Department of Electrical and\\Computer Engineering\\
%McGill University\\
%Montreal, Quebec}}
\author{Tianpei Chen, Harry Leib\\
 Department of Electrical and Computer Engineering\\
 McGill University\\
 Montreal, Quebec}

@@

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

\begin{document}
% make the title area
\begin{spacing}{2.5}
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal techniques for next generation wireless communication systems. Detection algorithms for large MIMO tend to require long simulation times and on conventional computer systems, In order to reduce the simulation times for decoders of large MIMO systems. In this paper we propose to use General Purpose Graphic Processing Units (GPGPU)to implement Fixed Complexity Sphere Decoder (FCSD) for the large scale MIMO system based on the dominant framework of GPGPU-Computed Unified Device Architecture (CUDA). The proposed CUDA-FCSD can chieve up to five times speedup over conventional CPU implementation while reaching the same Bit Error Rate (BER) performance. Furthermore the proposed algorithm demonstrates great potential to accelerate large MIMO systems.
\end{abstract}

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
%\IEEEpeerreviewmaketitle

\section{Introduction}
 \paragraph{} The present challenge in the wireless communication area is the sharp contrast between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (quality of service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) is thought to be a very promising technology that can help to solve this problem owing to several advantages it can offer. The core idea of MIMO is to use multiple antennas that at the transmitting and receiving sides that provide parallel data pipes in space resulting in spatial multiplexing gains\cite{oestges2010mimo}. For example the LTE standard supports up to 8 antennas at the base station\cite{dahlman20103g}. 
 \paragraph{} Large scale MIMO (also named massive MIMO) systems employ hundreds of low-power low-price antennas collocated at the base station and serving several tens of terminals at the same time\cite{rusek2013scaling}. This technology originated from multiuser MIMO (MU-MIMO) systems\cite{larsson2013massive}. Large scale MIMO will inherit and enhance all the advantages of conventional MIMO (spatial diversity gain and spatial multiplexing gain), as well as reduction of latency and improvement of system robustness. 
 A common decoder for MIMO system is the Sphere Decoder (SD) that perform as well as a Maximum Likelihood (ML) decoder with a dramatic complexity reduction\cite{viterbo1999universal}\cite{hassibi2005sphere}. There are two major problems with the traditional SD decoder: complexity depending on Signal to Noise Ratio (SNR), and sequential nature, which is contradictory with the requirement of a constant data processing throughput and parallel implementation for practical system. The Fixed Complexity Sphere Decoder (FCSD) overcomes these two disadvantages of the SD decoder by searching the solution along a fixed number of independent paths\cite{barbero2008fixing}.
\paragraph{}General purpose Graphic processing Units Computing (GPGPU) has become a common trend in industry and research in areas that need massive computations such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435} and finance\cite{grauer2013accelerating}. Compared to CPU, the GPU can naturally cope with computational intensive tasks, since most of the hardware resource are allocated for data processing and parallel computation, rather than caching and flow control. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. Computed unified Device Architecture (CUDA) is a programming model for GPGPU that was proposed by Nvidia and is becoming the dominant framework for GPGPU programming. It is based on popular programming languages(C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU implement parallel data processing, and has become a parallel computing platform for wide community of different applications.
\paragraph{}In this paper, a GPU based implementation of Fixed Complexity Sphere Decoder is proposed, A heterogeneous programming solution for FCSD preprocessing and blocked parallel path searching solution are presented, the speedup results of different influence factors are presented and analyzed, The proposed CUDA-FCSD implementation demonstrates a great potential for large MIMO systems.
\paragraph{}The rest of the paper is organized as follows. Section \ref{system} describes the MIMO systems model. Section \ref{programming model} introduces CUDA GPU architecture and the programming model. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are presented and analyzed. Finally conclusions are drawn in section \ref{conclusion}.    
\section{MIMO System Model}\label{system}
% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.8 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill December 27, 2012
\paragraph{}We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel, corresponding to the discrete time model:

@@

\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n}   \label{formula 1}

@@

$\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from the signal constellation $\mathbb{O}$ (4-QAM, 16-QAM, 64-QAM), of size $M$. The transmitted symbol vectors forms a $N_{t}$ complex signal vector space $\mathbb{O}^{N_{t}}$. $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. Furthermore,  $\mathbf{y}\in \mathbb{C}^{N_{r}\times 1}$ is the received symbol vector, $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identity distributed(i.i.d) circularly symmetric complex Gaussian zero mean components of unit variance, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) samples with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, $N_{0}$ denotes the average noise power spectrum density and $\frac{E_{s}}{N_{0}}$ denotes the signal to noise ratio (SNR). \\

\paragraph{} Assume we have the perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known at the receiver, as well as the SNR. The task of MIMO decoder is to estimate $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.

% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results.}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

\section{CUDA GPU Architecture and  Programming Model} \label{programming model}
\subsection{Memory Hierarchy}  
CUDA programming model is the C programming extension, with special defined syntax and structure. The serial code executes on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side, is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA, a block of cooperative threads form a thread block, in one block all the threads can share one pitch of on chip memory (shared memory), all the threads executed in one block can synchronize their action meaning the threads are permitted to run to the next instruction only when all the threads have reached the same instruction node. Blocks can be grouped into a one-dimensional, two-dimensional, or three-dimensional grid. Similarly, The threads can also be organized into one-dimensional, two-dimensional or three-dimensional blocks. The organization of the block and grid shape and dimension is determined by the application, Figure \ref{figure1} illustrates the 3 dimensional thread hierarchy of CUDA.
\begin{figure}[htb]
\centering
\includegraphics[scale=0.5]{CUDA_programming_hierachy.eps}
\caption{CUDA 3D thread hierachy\\ $5\times 5\times 5$ grid and each block contains $5\times 5\times 5$ threads\\}
\label{figure1}
\end{figure}

\subsection{Single Instruction Multiple Thread (SIMT) Model} Threads are scheduled, managed, created and executed in a group of 32 threads called warp, the programs executes on the device side are defined as $\mathit{kernel}$ function, when a $\mathit{kernel}$ function is invoked, the thread blocks are partitioned into warps, the threads in one warps executes the same instruction simultaneously, the different warps are scheduled by warp scheduler, when one warp is waiting for data loading the other warps begin to execute, this is called Single Instruction Multiple Thread (SIMT). This programming architecture is askin to Single Instruction Multiple Data (SIMD), however unlike the SIMD vectoring machines, SIMT can branch the behavior of each threads to achieve thread-level parallelism. The GPU use massive number of threads to hide the long latency of global memory instead of using large cache as CPU, in a word CPU cores are designed to minimize latency for each thread at a time, whereas GPUs are designed to handle a large number of concurrent, lightweight threads in order to maximize throughput.

>>>>

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[12pt,letter,final]{article}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
\usepackage[latin1]{inputenc}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage{graphicx}
%\usepackage[pdftex]{graphicx}
\usepackage{psfrag}
\usepackage{color}
\usepackage{epsfig}
\usepackage{appendix}
\usepackage{caption}
\usepackage{cite}
\usepackage{ifpdf}
\usepackage[cmex10]{amsmath}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{stfloats}
\usepackage{url}
\usepackage{fixltx2e}
\usepackage{setspace} 
\usepackage{diagbox}
\usepackage{subfigure}

@@

%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Report}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Tianpei Chen}
%\IEEEauthorblockA{ID: 25504946\\Department of Electrical and\\Computer Engineering\\
%McGill University\\
%Montreal, Quebec}}
\author{Tianpei Chen\\
 Department of Electrical and Computer Engineering\\
 McGill University}

@@

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}
\begin{document}
% make the title area
\begin{spacing}{2.5}
\maketitle
\section{Introduction}
Detection algorithm is one of the key components of Multiple-Input Multiple-Output (MIMO) systems. Designing of high performance and low complexity detector has become a bottleneck of Large MIMO systems. Recently, academia has found a property of large MIMO channel called channel hardening phenomenon \cite{hochwald2004multiple}\cite{tse2005fundamentals}\cite{marvcenko1967distribution}\cite{tulino2004random}, which states the phenomenon that with the number of receiving or transmitting antennas increasing, the variances of channel mutual information decrease. Simply saying, the channel "hardens". An useful aspect of channel hardening phenomenon is the off diagonal elements of $\mathbf{H}^{H}\mathbf{H}$ matrix becomes more and more ignorable comparing to diagonal elements when the number of receiving antennas or transmitting antennas becomes large.

 Channel hardening phenomenon provides an opportunity for designing computationally economical detection algorithms. For example, linear detectors (LDs) such as zero forcing (ZF) and minimum mean square error (MMSE) detectors need to calculate the inverse of $\mathbf{H}^{H}\mathbf{H}$ matrix. This process can be accelerated by approximate matrix inversion using series expansion techniques and deterministic approximations in large dimensions\cite{wu2013approximate}. Channel hardening phenomenon can also provide computational advances to machine learning based detectors, e.g. probability data association (PDA)\cite{fricke1900impact}\cite{pham2004generalized} and message passing detectors\cite{som2011low}\cite{goldberger2011mimo}\cite{lakshmi2013channel} and Monte-Carlo Markov Chain (MCMC) MIMO detectors\cite{farhang2006markov}\cite{datta2012novel}. This type of algorithms work well in large sparse systems.

In this report, we propose some useful insights of the orthogonal properties of channel. We define orthogonality measurement (om), which is a well round metric that considers both diagonal and off diagonal elements of $\mathbf{H}^{H}\mathbf{H}$.

The rest of the report is organized as follows, section \ref{system model} introduces system model. Section \ref{preliminary} provides some preliminaries of the definition of orthogonality measurement. Section \ref{log expectation of om} discusses the derivation of logarithmic expectation of orthogonality measurement. In section \ref{pdf of om} we obtain probability density function of orthogonality measurement. Some computer simulation results are presented in section \ref{computer simulation}.

\section{System Model}\label{system model}
We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel. Using a discrete time model, $\mathbf{y}\in\mathbb{C}^{N_{r}\times 1}$ is the received symbol vector written as:
\begin{equation}
\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n},   \label{formula 1}
\end{equation}
where $\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from a finite signal constellation alphabet $\mathbb{O}$ (e.g. 4-QAM, 16-QAM, 64-QAM) of size $M$. The possible transmitted symbol vectors $\mathbf{s}\in \mathbb{O}^{N_{t}}$, satisfy $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, where $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. Furthermore $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identically distributed (i.i.d) circularly symmetric complex Gaussian components of zero mean and unit variance. Finally, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) vector with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, where $N_{0}$ denotes the noise power spectrum density, and hence $\frac{E_{s}}{N_{0}}$ is the signal to noise ratio (SNR).

Assume the receiver has perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known, as well as the SNR. The task of the MIMO decoder is to recover $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.
\section{Preliminaries}\label{preliminary}
Orthogonality deficiency measures the how orthogonal a matrix is \cite{ma2008performance}, which is defined by
\begin{equation}
\phi_{od}=1-\frac{det(\mathbf{W})}{\prod_{i=1}^{N_{t}}||\mathbf{h}_{i}||^{2}},
\label{formula1}
\end{equation}
where $\mathbf{W}=\mathbf{H}^{H}\mathbf{H}$ denotes Wishart matrix, $\mathbf{h}_{i}$ denotes the $i$ th column of $\mathbf{H}$, $det(\cdot)$ denotes determinant operation, $||\cdot||^{2}$ denotes 2-norm operation.
In (\ref{formula1}), $||\mathbf{h}_{i}||^{2}=\sum_{i=1}^{N_{t}}|\mathbf{H}_{ij}|^{2}$, $\mathbf{H}_{ij}$ denotes the component of $\mathbf{H}$ at $i$ th row and $j$ th column. $\mathbf{H}_{ij}\sim Rayleigh(1/\sqrt{2})$, therefore $||\mathbf{h}_{i}||^{2}\sim Gamma(N_{r},1)$\cite{papoulis1996stochastic}. $Gamma(k,\theta)$ denotes Gamma distribution, with $k$ degrees of freedom. Furthermore, we have:
\begin{equation}
2||\mathbf{h}_{i}||^{2}\sim Gamma(N_{r},2)\sim\chi^{2}_{2N_{r}},
\label{formula2}
\end{equation} 
$\chi^{2}_{k}$ denotes chi-square distribution with $k$ degrees of freedom. For the sake of simplicity, (\ref{formula1}) can be changed to:
\begin{equation}
\phi_{om}=\frac{det(\mathbf{W})}{\prod_{i=1}^{N_{t}}||\mathbf{h}_{i}||^{2}}=\frac{2^{N_{t}}det(\mathbf{W})}{\prod_{i=1}^{N_{t}}2||\mathbf{h}_{i}||^{2}}.
\label{formula3}
\end{equation}
Taking logarithmic operation to $\phi_{om}$ we have 
\begin{equation}
\ln(\phi_{om})=N_{t}\ln(2)+\ln(det(\mathbf{W}))-\sum_{i=1}^{N_{t}}\ln(2||\mathbf{h}_{i}||^{2}),
\label{formula4}
\end{equation}
$\phi_{om}$ in (\ref{formula3}) is defined as Orthogonality Measure. Based on Hadamard's inequality ($\prod_{i=1}^{N_{t}}||\mathbf{h}_{i}||\geq det(\mathbf{H})$). $\phi_{om}\in [0,1]$. If $\phi_{om}$ is more closer to 1, $\mathbf{H}$ is closer to orthogonal matrix.

Because $\mathbf{W}=\mathbf{H}^{H}\mathbf{H}$, do QR factorization to $\mathbf{H}$ 
\begin{equation}
\mathbf{H}=\mathbf{Q}\mathbf{R},
\label{formula5} 
 \end{equation}
where $\mathbf{Q}\in\mathbb{C}^{N_{r}\times N_{t}}$ is a unitary matrix and $\mathbf{R}\in\mathbb{C}^{N_{t}\times N_{t}}$ is the upper triangular matrix. Using (\ref{formula5}), we have $\mathbf{W}=\mathbf{R}^{H}\mathbf{R}$. $r_{ii}$ denotes the $i$ th diagonal component of $\mathbf{R}$, thus $det(\mathbf{W})$ can be rewritten as:
\begin{equation}
det(\mathbf{W})=det(\mathbf{R}^{H}\mathbf{R})=det(\mathbf{R}^{H})det(\mathbf{R})\\=\prod_{i=1}^{N_{t}}r_{ii}^{H}\prod_{i=1}^{N_{t}}r_{ii}=\prod_{i=1}^{N_{t}}|r_{ii}|^{2}.
\label{formula6}
\end{equation}  
Notice that $\mathbf{R}$ can be viewed as the Cholesky factorization of $\mathbf{W}$. Therefore, we have
\begin{equation}
 ||\mathbf{h}_{i}||^{2}=\mathbf{W}_{ii}=\sum^{i-1}_{j=1}|r_{ji}|^{2}+|r_{ii}|^{2},
\label{formula7}
 \end{equation}
 where $\mathbf{W}_{ii}$ denotes the $i$ th diagonal element of $\mathbf{W}$.
 Thus based on (\ref{formula6}) and (\ref{formula7}), (\ref{formula3}) can be rewritten as:
\begin{equation}
\phi_{om}=\prod_{i=1}^{N_{t}}\frac{|r_{ii}|^{2}}{|r_{ii}|^{2}+\sum^{i-1}_{j=1}|r_{ji}|^{2}}.
\label{formula9}
\end{equation}  
\section{Logarithmic Expectation of Orthogonality Measurement}\label{log expectation of om}

Taking expectation of (\ref{formula4}), we have 
\begin{equation}
\mathbb{E}[\ln(\phi_{om})]=N_{t}\ln(2)+\mathbb{E}[\ln(det(\mathbf{W}))]-\sum_{i=1}^{N_{t}}\mathbb{E}[\ln(2||\mathbf{h}_{i}||^{2})].
\label{formula11}
\end{equation}
Consider $\mathbf{H}=[\mathbf{h}_{1}^{'},\mathbf{h}_{2}^{'},\cdots \mathbf{h}_{N_{r}}^{'}]^{'}$, where $\mathbf{h}_{i}$ denotes the $i$ th row of $\mathbf{H}$, because each component of $\mathbf{H}$ is mutually independent and subject to circularly symmetric complex Gaussian distribution, i.e. $h_{i}\sim \mathbb{C}N(\mathbf{0},\mathbf{I}_{N_{t}})$. Therefore, $\mathbf{W}=\mathbf{H}^{H}\mathbf{H}\sim \mathbb{C}W(N_{r}, \mathbf{I}_{N_{t}})$, $\mathbb{C}W(n,\mathbf{\Sigma})$ denotes complex Wishart distribution with $n$ degrees of freedom and covariance matrix $\mathbf{\Sigma}$. The logarithmic expectation of $\mathbf{W}$ can be rewritten as
\begin{equation}
\mathbb{E}[\ln(det(\mathbf{W}))]=\frac{\tilde{\Gamma}^{'}_{N_{t}}(N_{r})}{\tilde{\Gamma}_{N_{t}}(N_{r})}=\sum_{i=1}^{N_{t}}\psi(N_{r}-i+1),
\label{formula12}
\end{equation}
where $\tilde{\Gamma}_{m}(n)$ denotes the multivariate Gamma function. Proof: see Appendix A.

Because the logarithmic expectation of a Gamma distribution variable $\mathbb{g}\sim Gamma(n,\theta)$ can be written as:
\begin{equation}
\mathbb{E}[\ln(\mathbb{g})]=\psi(n)+\ln(\theta),
\end{equation}
where $\psi(n)$ denotes Digamma function. Thus according to (\ref{formula2}), we have:
\begin{equation}
\mathbb{E}[\ln(2||\mathbf{h}_{i}||^{2})]=\psi(N_{r})+\ln(2).
\label{formula14}
\end{equation}
Proof: see Appendix B
Based on (\ref{formula11})(\ref{formula12})(\ref{formula14}), The logarithmic expectation of $\phi_{om}$ can be written as:
\begin{eqnarray}
\nonumber
\mathbb{E}[\ln(\phi_{om})]&=&N_{t}\ln(2)+\sum_{i=1}^{N_{t}}\psi(N_{r}-i+1)-N_{t}\psi(N_{r})-N_{t}\ln(2)\\
&=& \sum_{i=1}^{N_{t}}\psi(N_{r}-i+1)-N_{t}\psi(N_{r})
\label{formula15}
\end{eqnarray}
\section{Probability Density Function of Orthogonality Measurement}\label{pdf of om}
Recall (\ref{formula9})
\begin{equation}
\phi_{om}=\prod_{i=1}^{N_{t}}\frac{|r_{ii}|^{2}}{|r_{ii}|^{2}+\sum_{j<i}|r_{ji}|^{2}}.
\end{equation}
All the components in $\mathbf{R}$ are independently distributed and $r_{ji}\sim \mathbb{C}N(0,1)$, $|r_{ii}|^{2}\sim Gamma((N_{r}-i+1),1)$ \cite{nagar2011expectations}. Because $|r_{ji}|\sim Rayleigh(1/\sqrt{2})$, $\sum_{j<i}|r_{ji}|^{2}\sim Gamma(i-1, 1)$. Defining $\alpha_{i}=\sum_{j<i}|r_{ji}|^{2}$ and $\beta_{i}=|r_{ii}|^{2}$, $\alpha_{i}$ and $\beta_{i}$ are mutually independent, therefore (\ref{formula9}) can be rewritten to 
\begin{equation}
\phi_{om}=\prod_{i=1}^{N_{t}}\frac{\beta_{i}}{\alpha_{i}+\beta_{i}},
\label{formula21}
\end{equation}
From \cite{gupta2004handbook}, if $X\sim Gamma(k_{1},\theta)$ and $Y\sim Gamma(k_{2},\theta)$, then $\frac{X}{X+Y}\sim B(k_{1},k_{2})$, where $B$ denotes Beta distribution. Therefore $\frac{\beta_{i}}{\beta_{i}+\alpha_{i}}\sim B(k^{i}_{1}, k^{i}_{2})$, where $k^{i}_{1}=N_{r}-i+1$, $k^{i}_{2}=i-1$. we define $\eta_{i}=\frac{\beta_{i}}{\beta_{i}+\alpha_{i}}$, it is obvious that $\eta_{i}$ are independently distributed. Based on (\ref{formula21}), we have 
\begin{equation}
\phi_{om}=\prod_{i=1}^{N_{t}}\eta_{i}.
\label{formula22}
\end{equation}
Therefore the density function of $\phi_{om}$ can be defined as
\begin{equation}
f_{\phi_{om}}(x)=\frac{1}{x}\sum_{\mathbf{j}}(\prod_{i=1}^{N_{t}}c(k_{1}^{i}k,k_{2}^{i},j^{i}))f(\ln(x)|\mathbf{k_{1}}+\mathbf{j}), 
\label{formula23}
\end{equation}
where $\sum_{\mathbf{j}}=\sum_{j^{1}}\sum_{j^{2}}\cdots \sum_{j^{N_{t}}}$, the range of $j^{i}\in [0, k_{2}^{i}-1]$, $c(k_{1}^{i}, k_{2}^{i}, j_{i})=(-1)^{j^{i}}$ $k_{2}^{i}-1\choose j^{i}$ $[(k_{1}^{i}+k_{2}^{i})\mathbb{B}(k_{1}^{i},k_{2}^{i})]^{-1}$, $\mathbb{B}(\alpha, \beta)$ denotes beta function. $f(\ln(x)|\mathbf{k_{1}}+\mathbf{j})=(\prod_{i=1}^{N_{t}}(k_{1}^{i}+j^{i}))\sum_{i=1}^{N_{t}}[exp((k_{1}^{i}+j^{i})\ln(x))/\prod_{j=1\\ j\neq i}^{N_{t}}(k_{1}^{j}+j^{j}-k_{1}^{i}-j^{i})]$. $\mathbf{k_{1}}+\mathbf{j}=[k_{1}^{1}+j^{1},\cdots k_{1}^{N_{t}-1}+j^{N_{t}-1},k_{1}^{N_{t}}+j^{N_{t}}]$. Proof: see Appendix C.

Consider logarithmic expectation of $\phi_{om}$, we have
\begin{equation}
E[\ln(\phi_{om})]=\sum_{i=1}^{N_{t}}E[\ln(\eta_{i})],
\label{formula24}
\end{equation}
where $E[\ln(\eta_{i})]=\psi(k^{i}_{1})-\psi(k^{i}_{1}+k^{i}_{2})$, thus we have 
\begin{equation}
E[\ln(\phi_{om})]=\sum_{i=1}^{N_{t}}\psi(N_{r}-i+1)-N_{t}\psi(N_{r}).
\label{formula25}
\end{equation}
we can find (\ref{formula25}) is consistent with (\ref{formula15}).

\section{Computer Simulations}\label{computer simulation}
 Computer simulations are made for different sizes of V-BLAST MIMO systems, with $5\leq N_{r} \leq 100, 5 \leq N_{t} \leq N_{r}$, the empirical estimation of logarithmic expectation of $\phi_{om}$, $E[\ln(\phi_{om})]_{em}$, is calculated by taking average over $1e4$ channel realizations for each size of MIMO systems, as shown in Fig. \ref{figure1}, the Theoretical logarithmic expectation of $\phi_{om}$ $E[\ln(\phi_{om})]_{t}$ in (\ref{formula25}) is plotted in Fig. \ref{figure2}. Average deviation between $E[\ln(\phi_{om})]_{em}$ and $E[\ln(\phi_{om})]_{t}$ is also calculated, $V_{em-t}= 7.3043e-04$.

\begin{figure}[htb]
\centering
\includegraphics[scale=0.6]{NtE.eps}
\caption{Relation between $N_{t}$ and $E[\ln(\phi_{om})]_{t}$}
\label{figure3}
\end{figure}

Fig. \ref{figure3} demonstrates the relation between the number of users ($N_{t}$) and $E[\ln(\phi_{om})]_{t}$ under cases of different numbers of antennas at base station ($N_{r}$). From Fig. \ref{figure3}, we can see, on the one hand, with $N_{r}$ fixed, $E[\ln(\phi_{om})]$ decreases while $N_{t}$ increases, however the gradient of each curve becomes more and more gentle. On the other hand, when $N_{r}$ becomes larger $E[\ln(\phi_{om})]$ becomes more insensitive to variation of $N_{t}$.
\begin{figure}[htb]
\centering
\centering\includegraphics[width=0.7\textwidth, height=5cm]{LogE_om_em.eps}
\caption{Empirical Estimation $E[\ln(\phi_{om})]_{em}$}
\label{figure1}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.7\textwidth, height=5cm]{LogE_om_t.eps}
\caption{Theoretical $E[\ln(\phi_{om})]_{t}$}
\label{figure2}
\end{figure}

\begin{appendices}
\section{Appendix A}
Let $\mathbf{A}\in \mathbb{C}^{m\times m}$, $A\sim \mathbb{C}W(n, \mathbf{\Sigma})$, $\mathbb{C}W(n, \mathbf{\Sigma})$ denotes complex Wishart distribution with $n$ degrees of freedom and covariance matrix $\mathbf{\Sigma}$. It is obvious $\mathbf{A}$ is Hermition positive definite matrix, $\mathbf{A}=\mathbf{A}^{H}>0$.

The p.d.f function of $\mathbf{A}$ can be written as\cite{nagar2011expectations}:
\begin{equation}
f(A)=\{\tilde{\Gamma}_{m}(n)det(\mathbf{\Sigma})^{n} \}^{-1}det(A)^{n-m}etr(-\mathbf{\Sigma}^{-1}\mathbf{A}),
\label{Appendequa1}
\end{equation}
where $\tilde{\Gamma}_{m}(\beta)$ denotes multivariate complex Gamma function defined by:
\begin{equation}
\tilde{\Gamma}_{m}(\beta)=\pi^{\frac{m(m-1)}{2}}\prod_{i=1}^{m}\Gamma(\beta-i+1)\quad Re(\beta)>m-1.
\label{Appendequa2}
\end{equation}
Furthermore,from \cite{nagar2011expectations}, we have

@@

\tilde{\Gamma}_{m}(\beta)=\int_{\mathbf{X}=\mathbf{X}^{H}>0}etr(-\mathbf{X})det(\mathbf{X})^{\beta-m}d
\mathbf{X} \quad Re(\beta)>m-1.
\label{Appendequa3}

@@

We derive logarithmic expectation of $det(\mathbf{A})$
